Tracing the Hercules stream around the Galaxy

### 2010-09-12 tex/response.txt

	 The paper titled Tracing the Hercules stream around the Galaxy by Jo
	 Bovy is concerned with the origin and detectability of this
	 elusive stream. Overall the paper is well written and
	 clear. I have a few major concerns and some minor concerns
	 that should be addressed before the paper can be accepted for
	 publication.

	 Major concerns:

	 1. Sect 4, page 6, last paragraph: The author discusses in
	 some detail, and shows plots to illustrate, the possible
	 implications of e.g.  errors in the distance determination
	 etc. The one error that is not address explicitly in the text
	 or plots is the error on the measured radial velocities - how
	 well do we need to determine these velocities in order to be
	 able to unambiguously detect the Hercules stream? 1, 5, or 10
	 km/s? Or some other number? And what happens when V0 is
	 absolutely known or if it also has an error? Perhaps the
	 result is that the errors do no matter, but in that case this
	 should be shown and arguments to why it does not matter
	 should be given. Also, will either the RVS on board Gaia or
	 APOGEE deliver radial velocities with small enough errors to
	 detect the Hercules stream?

I added a panel to Figure 5 showing the influence of realistic
line-of-sight velocity uncertainties. The expected uncertainties for
APOGEE and HERMES are below 1 km/s; for such small uncertainties the
uncertainty-free prediction and the uncertainty-convolved prediction
are indistinguishable.

I added explicit references to the Gaia, APOGEE, and HERMES
uncertainties in the sections describing these surveys.

BOVY: V0

@@

	 3. I find Discussion in Sect 5 (for both Gaia and APOGEE)
	 somewhat sketchy. For example, no mention of the fact that
	 Gaia will provide radial velocities for stars down to roughly
	 17 magnitude is mentioned.  Should the reader conclude that
	 Hercules could be detected with the tangential velocities
	 alone? In that case, how does that work with the tests
	 presented in this article? Again, here the intrinsic
	 brightness of the tracers is of some interest.

I added a reference to the line-of-sight velocity accuracies expected
for Gaia, and an explicit comment that the velocities of giants will
be measured accurately enough to verify the predictions of this paper.

	 4. The discussion could in general be deeper, and also connect to
	 the abstract (see below).

I hope that by adding the general discussion and the greater detail in
discussing the specific surveys that could detect the Hercules
feature, the discussion now has the required depth.

@@

	 1. On page 2, last paragraph of the introduction it is stated
	 that "propose and ... test of the bar resonance model for the
	 Hercules stream". However, this statement is not really
	 carried through in the text. It might be a good idea to
	 connect back to this statement in e.g. Sect 5 and tell the
	 reader how "testable" this model is once all the
	 uncertainties of data and shortcomings of the missions has
	 been discussed. The test and properties of the bar are also
	 mentioned in the abstract, however, there is no discussion in
	 the paper of how the properties of the bar are constrained by
	 the proposed observational test. Add this discussion, which
	 would be very interesting.

	 2. Page 2, statement at the end of the first paragraph "..up
	 to the task of separating out these effects." This statement
	 should be supported with a few relevant references.

	 3. On page 5, sect 4, third paragraph, last sentence. I do
	 not understand what the last sentence means - could the
	 author please re-phrase it? The sentence somehow, does not
	 connect to the previous text.

I thought this followed clearly from the sentences before it, but I
have changed this sentence to:

### 2010-09-12 tex/hercules.tex
groups-simulations: note about APOGEE
hercules: add general discussion

\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath}
%\usepackage{color,hyperref}
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\figurenames}{\figurename s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\tvector}[1]{\boldsymbol{\vec{#1}}}
\newcommand{\vx}{\tvector{x}}
\newcommand{\vv}{\tvector{v}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\Gaia}{\emph{Gaia}}
\newcommand{\apogee}{APOGEE}
\newcommand{\hipparcos}{\emph{Hipparcos}}
\newcommand{\vR}{\ensuremath{v_R}}
\newcommand{\vphi}{\ensuremath{v_{\phi}}}
\newcommand{\vo}{\ensuremath{v_0}}
\newcommand{\Ro}{\ensuremath{R_0}}
\newcommand{\ro}{\Ro}
\newcommand{\Ab}{\ensuremath{A_b}}
\newcommand{\Rb}{\ensuremath{R_b}}
\newcommand{\Omegab}{\ensuremath{\Omega_b}}
\newcommand{\fdehnen}{\ensuremath{f_{\text{Dehnen}}}}
\newcommand{\sigmaR}{\ensuremath{\sigma_R}}
\newcommand{\rE}{\ensuremath{R_e}}
\newcommand{\Lc}{\ensuremath{L_c}}
\newcommand{\Rs}{\ensuremath{R_s}}
\newcommand{\Rsigma}{\ensuremath{R_{\sigma}}}
\newcommand{\Rolr}{\ensuremath{R_{\text{OLR}}}}

@@

\begin{abstract}
It has been proposed that the Hercules stream, a group of co-moving
stars in the Solar neighborhood offset from the bulk of the velocity
distribution, is the result of resonant interactions between stars in
the outer disk and the Galactic bar. So far it has only been seen in
the immediate Solar neighborhood, but the resonance model makes a
prediction over a large fraction of the Galactic disk. I predict the
distribution of stellar velocities and the changing Hercules feature
in this distribution as a function of location in the Galactic disk in
a simple model for the Galaxy and the bar that produces the observed
Hercules stream. The Hercules feature is expected to be strong enough
to be unambiguously detected in the distribution of line-of-sight
velocities in selected directions. I identify quantitatively the most
promising lines of sight for detection in line-of-sight velocities
using the Kullback-Leibler divergence between the predictions of the
resonance model and an axisymmetric model; these directions are at
$250^{\circ} \lesssim l \lesssim 290^{\circ}$. The predictions
presented here are only weakly affected by distance uncertainties,
assumptions about the distribution function in the stellar disk, and
the details of the Galactic potential including the effect of spiral
structure. \Gaia\ and future spectroscopic surveys of the Galactic
disk such as \apogee\ will be able to robustly test the origin of the
Hercules stream and constrain the properties of the Galactic bar.
\end{abstract}

@@

Finally, in \figurename~\ref{fig:1dvar} I show the influence of
various nuisance and dynamical parameters on the predicted
line-of-sight velocity distribution for the most promising region that
is more than 3 kpc away. We see in this \figurename\ that the effect
of distance uncertainties is negligible. The exact shape and
parameters of the stellar distribution function also only minimally
affect the predicted line-of-sight velocity distribution. Changing the
scale-length within its current uncertainty changes the prediction
only in the amplitude of the effect: The Hercules stream is more
prominent for shorter scale-lengths since then there are relatively
more stars in the inner Galaxy that are brought to the Solar circle by
the interaction with the bar. Using a Shu distribution function
\citep{shu69a} instead of the Dehnen distribution function also gives
very similar predictions. Changing the axisymmetric potential by
adjusting the shape of the rotation curve shifts the peak as expected
\citep[see][]{dehnen00a}, but the Hercules feature nevertheless
remains detectable.

@@

@@

@@

@@

\bibitem[Sellwood(2010)]{sellwood10a}
  Sellwood,~J.~A., 2010,
  \mnras, submitted, 
  arXiv:1001.5197v1 [astro-ph.GA]

@@

\bibitem[Shen \etal(2010)]{Shen10a}
  Shen,~J., \etal, 2010, \apj L, submitted, arXiv:1005.0385v1 [astro-ph.CO]

@@

\clearpage
\begin{figure}
\includegraphics[width=0.35\textwidth]{distuncertain.ps}
\includegraphics[width=0.35\textwidth]{vlosuncertain.ps}\\
\includegraphics[width=0.35\textwidth]{df.ps}
\includegraphics[width=0.35\textwidth]{slope.ps}\\
\includegraphics[width=0.35\textwidth]{rolr.ps}
\includegraphics[width=0.35\textwidth]{barstrength.ps}
\caption{Variation of the predicted velocity distribution at R = 1.075
  \Ro , $\phi = -21^{\circ}$ (d= 0.4 \Ro, $l$ = 270$^{\circ}$) with
  distance uncertainties; line-of-sight velocity uncertainties;
  parameters and shape of the distribution function; shape of the
  rotation curve; and bar strength.}\label{fig:1dvar}
\end{figure}

>>>>

\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath}
%\usepackage{color,hyperref}
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\figurenames}{\figurename s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\tvector}[1]{\boldsymbol{\vec{#1}}}
\newcommand{\vx}{\tvector{x}}
\newcommand{\vv}{\tvector{v}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\Gaia}{\emph{Gaia}}
\newcommand{\apogee}{APOGEE}
\newcommand{\hermes}{HERMES}
\newcommand{\hipparcos}{\emph{Hipparcos}}
\newcommand{\vR}{\ensuremath{v_R}}
\newcommand{\vphi}{\ensuremath{v_{\phi}}}
\newcommand{\vo}{\ensuremath{v_0}}
\newcommand{\Ro}{\ensuremath{R_0}}
\newcommand{\ro}{\Ro}
\newcommand{\Ab}{\ensuremath{A_b}}
\newcommand{\Rb}{\ensuremath{R_b}}
\newcommand{\Omegab}{\ensuremath{\Omega_b}}
\newcommand{\fdehnen}{\ensuremath{f_{\text{Dehnen}}}}
\newcommand{\sigmaR}{\ensuremath{\sigma_R}}
\newcommand{\rE}{\ensuremath{R_e}}
\newcommand{\Lc}{\ensuremath{L_c}}
\newcommand{\Rs}{\ensuremath{R_s}}
\newcommand{\Rsigma}{\ensuremath{R_{\sigma}}}
\newcommand{\Rolr}{\ensuremath{R_{\text{OLR}}}}

@@

\begin{abstract}
It has been proposed that the Hercules stream, a group of co-moving
stars in the Solar neighborhood offset from the bulk of the velocity
distribution, is the result of resonant interactions between stars in
the outer disk and the Galactic bar. So far it has only been seen in
the immediate Solar neighborhood, but the resonance model makes a
prediction over a large fraction of the Galactic disk. I predict the
distribution of stellar velocities and the changing Hercules feature
in this distribution as a function of location in the Galactic disk in
a simple model for the Galaxy and the bar that produces the observed
Hercules stream. The Hercules feature is expected to be strong enough
to be unambiguously detected in the distribution of line-of-sight
velocities in selected directions. I identify quantitatively the most
promising lines of sight for detection in line-of-sight velocities
using the Kullback-Leibler divergence between the predictions of the
resonance model and an axisymmetric model; these directions are at
$250^{\circ} \lesssim l \lesssim 290^{\circ}$. The predictions
presented here are only weakly affected by distance uncertainties,
assumptions about the distribution function in the stellar disk, and
the details of the Galactic potential including the effect of spiral
structure. \Gaia\ and future spectroscopic surveys of the Galactic
disk such as \apogee\ and \hermes\ will be able to robustly test the
origin of the Hercules stream and constrain the properties of the
Galactic bar.
\end{abstract}

@@

Finally, in \figurename~\ref{fig:1dvar} I show the influence of
various nuisance and dynamical parameters on the predicted
line-of-sight velocity distribution for the most promising region that
is more than 3 kpc away. We see in this \figurename\ that the effect
of distance and realistic line-of-sight velocity uncertainties is
negligible. The exact shape and parameters of the stellar distribution
function also only minimally affect the predicted line-of-sight
velocity distribution. Changing the scale-length within its current
uncertainty changes the prediction only in the amplitude of the
effect: The Hercules stream is more prominent for shorter
scale-lengths since then there are relatively more stars in the inner
Galaxy that are brought to the Solar circle by the interaction with
the bar. Using a Shu distribution function \citep{shu69a} instead of
the Dehnen distribution function also gives very similar
predictions. Changing the axisymmetric potential by adjusting the
shape of the rotation curve shifts the peak as expected
\citep[see][]{dehnen00a}, but the Hercules feature nevertheless
remains detectable. The properties of the bar itself, \eg, its pattern
speed through \Rolr\ or its strength $\alpha$, change the predicted
pattern, but they do not significantly alter the prospects for
detection.

@@

The discussion in the previous sections shows that the bar-origin
paradigm for the Hercules moving group makes strong predictions about
how the Hercules feature should show up when we probe the velocity
distribution in parts of the Galaxy far removed from the Solar
neighborhood. In order to most strongly test the bar-origin for the
Hercules moving group it is best to extrapolate this picture far from
the Solar neighborhood and look at regions further than a few kpc
removed. \Gaia 's parallaxes, proper motions, and mission-averaged
line-of-sight velocities for dwarfs nearby and giants at distances
greater than a few kpc will allow the full two-dimensional velocity
distribution to be confronted with the predictions of
\figurename~\ref{fig:rphi2d} (see below). However, the final
\Gaia\ catalog will only be available around 2020.

The predicted line-of-sight velocity distributions in the previous
section show that the bar-origin theory for the Hercules moving group
can be tested with line-of-sight velocities alone. At distances
greater than a few kpc this will mean observing giants to trace the
velocity distribution. Large, planned spectroscopic surveys with
line-of-sight velocity uncertainties less than 1 km s$^{-1}$ can test
the predictions made in this paper in the next few years (see
below). These surveys will not need to observe many stars to test the
predictions laid out here: Drawing samples from the predicted velocity
distribution in \figurename~\ref{fig:1dvar} shows that about 500 stars
suffice to distinguish the bar-origin model from the axisymmetric
prediction (based on a KS test with $p < 0.01$). Since large distance
uncertainties do not affect the prediction very much (see
\figurename~\ref{fig:1dvar}), these stars can even be sampled
throughout a volume of 500 pc or more.

The bottom two panels of \figurename~\ref{fig:1dvar} show that the
predicted line-of-sight velocity distribution depends strongly on the
dynamical parameters of the bar. Changing the pattern speed of the bar
through \Rolr\ shifts the predicted distribution and in particular the
location of the Hercules feature. While this shift at a particular
location is degenerate with the shift due to the uncertainty in the
local circular velocity, results from various locations can be
combined to break this degeneracy and the local circular velocity is
also already quite well constrained \citep[\eg,][]{Bovy09b}. Changing
the strength of the bar changes the relative heights of the two main
peaks in the predicted velocity distribution. However, the main peak
of the velocity distribution might not be as simple as the one
predicted here due to other dynamical effects. Integrated measures,
such as the number of stars in the main peak and the number of stars
in the Hercules feature, could mitigate this somewhat. While
predicting actual constraints would involve detailed simulations of
the expected data, it is clear that constraints on the bar can be
derived from the experiment proposed in this paper.

@@

BOVY: 500 stars is enough for APOGEE to distinguish from axisymmetric
background.

@@

\bibitem[Bovy, Hogg, \& Rix(2009)]{Bovy09b}
  Bovy,~J., Hogg,~D.~W., \& Rix, H.-W., 2009,
  \apj, 704, 1704

@@

\bibitem[Sellwood(2010)]{sellwood10a}
  Sellwood,~J.~A., 2010,
  \mnras, in press,
  arXiv:1001.5197v1 [astro-ph.GA]

@@

\bibitem[Shen \etal(2010)]{Shen10a}
  Shen,~J., \etal, 2010, \apj, 720L, 72

@@

\clearpage
\begin{figure}
\includegraphics[width=0.35\textwidth]{distuncertain.ps}
\includegraphics[width=0.35\textwidth]{vlosuncertain.ps}\\
\includegraphics[width=0.35\textwidth]{df.ps}
\includegraphics[width=0.35\textwidth]{slope.ps}\\
\includegraphics[width=0.35\textwidth]{rolr.ps}
\includegraphics[width=0.35\textwidth]{barstrength.ps}
\caption{Variation of the predicted velocity distribution at R = 1.075
  \Ro , $\phi = -21^{\circ}$ (d= 0.4 \Ro, $l$ = 270$^{\circ}$) with
  distance uncertainties; line-of-sight velocity uncertainties;
  parameters and shape of the distribution function; shape of the
  rotation curve; bar \Rolr; and bar strength.}\label{fig:1dvar}
\end{figure}
